GeckoGraph: A Visual Language for Polymorphic Types

### 2025-10-24 main.tex
Updates

%% Abstract
\begin{abstract}
Polymorphic types are an important feature in most strongly typed programming languages. They allow functions to be written in a way that can be used with different data types, while still enforcing the relationship and constraints between the values. However, programmers often find polymorphic types difficult to use and understand, and tend to reason with a mindset of concrete types. We propose GeckoGraph, a graphical notation for types. GeckoGraph aims to accompany traditional text-based type notation and to make reading, understanding, and comparing types easier. We conducted a large-scale human study using GeckoGraph compared to text-based type notation. To our knowledge, this is the largest controlled user study on functional programming ever conducted. Our results show that no significant difference was found in time and success rate between the programmers who use GeckoGraph and those who do not. However, we also find that for the most challenging task of the user study, GeckoGraph improves the success rate by 14.5\% for beginners and 11.2\% for less experienced users, suggesting the current iteration of GeckoGraph may be more useful as a teaching and learning tool. 
\end{abstract}

@@

 We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern Human-Computer Interaction (HCI) techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. We designed a gamified user study \textit{ZeroToHero}\footnote{Available at https://zerotohero.fly.dev}, evaluating 714 participants' performance to solve a series of programming puzzles using GeckoGraph versus text-based type annotation, which is, to our knowledge, the largest controlled user study of a functional programming tool or concept.  Overall, no significant difference was found in either time or success rate between the programmers who use GeckoGraph and those who do not. However, we find that in the hardest task of the user study, GeckoGraph improves the success rate by 14.5\%  for novice users (less than 1 year of experience) and 11.2\% for less experienced users (1-3 years of experience), suggesting the current iteration of GeckoGraph may be useful as a teaching and learning tool.

@@

\subsection{Structural Type Systems}

Unlike the nominal type systems found predominantly in languages such as Java and C \cite{AbdelGawad2013-hh}, many functional languages heavily rely on structural typing. In nominal systems, two types are considered distinct even if they have identical structure but different names. In contrast, structural type systems consider types equivalent as long as their structures match, regardless of their names. This approach promotes flexibility, as types can be defined inline without introducing new names, and values are automatically valid for a type when they conform to the structure.  However, this lack of nominality can also make complex types harder to read and reason about. For instance, when encountering type errors, programmers can face two unmatching types with a complex structure, and understanding where exactly the root error is becomes a tedious task of string comparison.

\subsection{Algebraic Data Types}

Algebraic data types (ADTs) \cite{Milner1978-qz} allow developers to construct complex types from simpler ones using sum and product operations. A {\bf sum type}, often called a tagged union, $A + B$, represents a type that can be a value of A or a value of B. Conversely, a {\bf product type} $A \times B$ represents a tuple type composed of two values of both {\tt A} and {\tt B}. This formalization allows precise and composable type construction for common data structures such as lists, tuples, and result types.
For example, in Haskell, types like \texttt{data Either a b = Left a | Right b} and \texttt{data Maybe a = Nothing | Just a} provide basic building blocks for error handling, and the type \texttt{data List a = [] | a : List a } provides a succinct recursive definition for lists. This illustrates the expressive power of ADTs. However, in practice, type definitions can grow to be complex structures, for instance \texttt {Maybe (Either [Maybe a] (Either a String))} as programs grow, causing difficulties to interpret and maintain.

\subsection{Parametric Polymorphism}

Parametric polymorphism (often called generics) allows functions and data structures to operate uniformly across different types. For instance, in Haskell, the function \texttt{head :: [a] -> a}, taking a list as input and returning the first element, is a polymorphic function that works on lists of any type; integers, strings, and functions are all valid candidates for the generic variable \texttt{a}.  It provides a strong and principled form of abstraction, enabling code reuse without sacrificing type safety.

@@

\subsection{Type Inference}

Type inference, or implicit typing, reduces the need for explicit type annotations by allowing the compiler to deduce types automatically.
Most modern languages employ some level of inference. For instance, Java’s var keyword (introduced in version 10) infers variable types at compile time based on their assigned values \cite{Java_Developers2023-an}, saving programmers from writing redundant type declarations, such as \textit {Car car = new Car()}.

@@

To summarize, the type systems of functional programming languages provide rigorous tools for specifying and enforcing program behavior. By encoding system constraints directly in types, they prevent many classes of runtime errors. However, this rigor introduces trade-offs: a steeper learning curve and less flexibility for exploratory or prototype development. In the following sections, we explore how GeckoGraph supports these type-system features by visualizing type-level information that is often implicit or difficult for humans to interpret.

@@

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, a type-class gliph is embedded in every type variable that requires this class. This means when displaying the type \texttt{(==) :: Eq a => 
a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in twice on both occurrences of the type variable \texttt{a}.

@@

\begin{figure}[]
  \includegraphics[width=0.6\linewidth]{figures/Add3}
  \caption{\label{fig:add3} An implementation of function \texttt{add3} but the author missed an (+) from the correct implementation (.) ((+) .) (+). GeckGraph highlights the difference in arity and reveals that a binary function is needed on the right-hand side for the arity to match. }
\end{figure}

@@

\paragraph{Provide Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references, or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail, a sepecific query that is hard to execute with point and click over text based type annotation.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/Advanced}
  \caption{\label{fig:advanced}
  Advanced features of GeckoGraph. (A) GeckoGraph supports Kind Visualization if the inferred kind is greater than \texttt{*}. (B) GeckoGraph supports qualified constraints by extending the extended area across multiple type variables. (C) GeckoGraph supports Multiple Parameter Type Classes, using different shapes of the same color to indicate that multiple variables must satisfy certain type classes collectively. (D) GeckoGraph supports the precise selection of its sub-structures. }
\end{figure}

\section{Evaluation} \label{sec:evaluation}
To evaluate the usefulness of GeckoGraph, we designed and conducted a gamified controlled experiment. The study, "Zero to Hero", contains 10 tasks (levels) of varying difficulty. At each level, participants are asked to implement a function named "\texttt{zeroToHero}" with the help of a limited list of provided functions. These functions are different at each level, and the target types of Zero and Hero vary at each level as well. The details of each level are provided in the Appendix \ref{levels}.

@@

We illustrate the tasks of the user study using levels 2 and 3 as examples. For level 2 (Fig. \ref{fig:level-example}), the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero b}. The available functions are \texttt{runZero::Zero a -> a}, \texttt{makeHero:: a -> Hero a}. A generic function \texttt{(\$)} is provided to improve the ergonomics of composing functions, but all tasks can be completed without the use of generic functions. The possible solution and other details of the level can be found in Appendix \ref{levels}. To complete this level, players only need to use each of the provided functions once. One possible solution to this task would be {\tt zeroToHero z = makeHero (runZero z)}. The user illustrated in Fig. \ref{fig:level-example} did not reach a correct solution, as the type signature and GeckoGraphs in {\it To Implement} and {\it Your Result} do not match.

@@

In level 3 (Fig. \ref{fig:level3}),  players need to implement the function {\tt zeroToHero :: Zero a -> Hero (a, a)}.  In this level, not all provided functions are used in the final result. In particular,  players must discover that only \texttt{f1} and \texttt{f3} are necessary to produce the desired results. One implementation that satisfies the target type is \texttt{zeroTohero z = f3 (f1 z)}. Another possible implementation is {\tt zeroToHero z = f3 . f1 \$ z}, exercising the Haskell function composition idioms. Again, the user illustrated in Fig. \ref{fig:level3} did not reach a correct solution, as the type signature and GeckoGraphs in {\it To Implement} and {\it Your Result} do not match.

@@

The experiment uses a between-subjects design. Participants are assigned to one of two groups. The tasks assigned to the two groups are in the same order. Participants receive both treatments (with and without GeckoGraph) during their runs, but in a different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same, but with the order flipped. The number of participants in the two groups is counterbalanced.

@@

\subsection{Task Design} \label{subsection:task}
During the study, participants receive 10 tasks, represented by the 10 levels of the user study. The tasks start off easy but gradually increase in difficulty.  In each task, a target type of the function \texttt{zeroToHero} is shown to the participants. Participants are provided with a list of available functions to implement the target function and to ensure it assumes the target type. This is to simulate the tasks of selecting useful functions from a library. Participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

\subsection{Measurements}
During the study, the time spent by participants on each task is recorded. We also record the resulting status of each level, whether it is a success or a failure. Before starting the study, participants nominate their level of Haskell experience on a four-level scale: beginner, familiar,  knowledgeable, and expert.  If a participant has completed all 10 levels (with the help of skipping), we invite the participant to complete a post-study survey. In it, we ask for their opinion on how intuitive the GeckoGraph design is, how distracting they find GeckoGraph, and how helpful GeckoGraph is, using a seven-point scale. In the end, we ask a few open-ended questions, inviting participants to provide their experience using GeckoGraph and their expectations about the potential applications of GeckoGraph.

Data collection from the study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero game is open source and available for free evaluation\footnote{https://zerotohero.fly.dev}, and repeating our experiment, but no further data was collected.

@@

\subsection{Success rate}
We saw that, overall, GeckoGraph provides a higher success rate (96.88\%) than text-based type notation (94.62\%). This trend can be seen in every experienced group: beginner group (95.12\% vs. 92.68\%), familiar group (97.39\% vs. 93.34\%), knowledgeable group (96.82\% vs.96.06\%), and expert group (98.2\% vs.96.40\%). We saw the significance decrease as the user's experience increased. When performing a proportion test on each group, we see that the effect is most significant with the beginner group and reject the null hypothesis (z score = 2.0228, p-value = 0.0431), followed by the familiar group (z score = 1.7495, p-value = 0.0802). The knowledgeable group (z score = 1.0295, p-value = 0.3032) and the expert group (z score = 0.8660, p-value = 0.3756) show less significant differences between treatments.

When breaking down the result in each task (Fig. \ref{fig:success-rate}), we were able to reject the null hypothesis in task 10 of the beginner group and task 10 of the familiar group (Fig. \ref{fig:success-rate}). We will address this correlation in Section~\ref{sec:discussion}.

Level 10 is the most cognitively demanding: the provided functions are the lengthiest, and the solution is challenging and deceptive. Programmers must use the provided rotate function {\it f2} to adjust the positions of the four components of the {\it Zero} value four times to get to the correct position. Completing the level requires applying the functions {\it f1}, {\it f2}, and {\it f3} in a nested sequence that is 8 layers deep. Consequently, programmers need to frequently switch contexts to recall the purpose of each function. We speculate that this level demands the mental flexibility typical of a Haskell expert to internalize these functions quickly. GeckoGraph helps less experienced programmers to achieve the same level of comprehension fluency by adding structural clarity.

@@

\subsection{Qualitative Feedback}
In their responses to the post-study survey, most programmers believe that the design of GeckoGraph is intuitive and that its appearance in the interface does not cause distraction.
For the question ``Do you find the GeckoGraph distracting", most of the participants rated a negative score, with an average of 2.88 (Fig. \ref{fig:qualitative} left). For the question ``How intuitive do you find the GeckoGraph?", we saw a reverse correlation of experience (Fig. \ref{fig:qualitative} Middle): experts find the GeckoGraph most intuitive (5.07), followed by the knowledgeable group (4.87), and the familiar group (4.80). The beginner group found it to be the least intuitive but still rated a positive score of 4.71. 
When answering the question ``How helpful do you find GeckoGraph in finding the solution?", the answer is more divided into different experience groups (Fig. \ref{fig:qualitative} right). It is slightly positive for beginners (mean=4.25) and slightly negative for the other groups, the familiar group (mean=3.86) and the knowledgeable group (mean=3.32). The expert group finds that GeckoGraph is the least helpful (mean=2.95)

Two interesting trends emerge. First, the more experienced the users are, the more likely they are to find GeckoGraph intuitive and easy to use. This aligns with our design decision to make GeckoGraph resemble text-based type annotation. Those familiar with text-based type annotation find GeckoGraph second nature to read. Second, the correlation reverses when it comes to perceived helpfulness. The more experienced the users are, the less likely they are to find GeckoGraph helpful. We speculate this pattern occurs because users who are experts in text-based type annotations naturally prefer reading text-based notation and tend to ignore GeckoGraph, diminishing its usefulness. Conversely, novice users are more likely to be drawn to GeckoGraph and find that its visualization offers more clarity and useful insights.

@@

\paragraph{Task design}
In our human study, most of the provided functions are very abstract. These functions are created by the authors solely for the gamified study. They are designed to be different from well-known Haskell functions to minimize the familiarity variable. They are also designed with an interest in being challenging and fun. These functions may not be the most representative of real-world Haskell programming.

@@

\subsection{Strengths}\label{sec:discussion:strengths}
Overall, our experimental results show a statistically significant difference in task completion time between conditions with and without GeckoGraph. However, we find using GeckoGraph significantly improves task success rates in the hardest task (task 10) across the two less experienced user groups by 14.5\% and 11.2\%, respectively. 
To interpret the observed advantages, we discuss the practical benefits that GeckoGraph offers to programmers.

@@

\section{Conclusion and Future Work}
In this paper, we propose GeckoGraph, a graphical notation for type annotations in functional programming languages. GeckoGraph aims to accompany traditional text-based type notation and to make reading, understanding, and comparing types easier. We conducted a large-scale human study using GeckoGraph compared to text-based type notation, the largest user study on functional programming we are aware of. Although no significant effect of GeckoGraph was found on the overall time and success rate, we find that GeckoGraph significantly improves the success rate for less experienced users in tackling more difficult tasks.

>>>>

%% Abstract
\begin{abstract}
Polymorphic type is an important feature in most strongly typed programming languages. They allow functions to be written in a way that can be used with different data types, while still enforcing the relationship and constraints between the values. However, programmers often find polymorphic types difficult to use and understand, and tend to reason with a mindset of concrete types. We propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation by making reading, understanding, and comparing types easier. We conducted a large-scale user study using GeckoGraph, comparing it to text-based type notation. To our knowledge, this is the largest controlled user study on functional programming ever conducted. 
The results show no overall significant differences in task completion time or success rate. However, we observed a modest improvement in success rate among beginner participants. For the most challenging task, GeckoGraph improved success rates by 14.5\% for beginners and 11.2\% for less experienced users. These findings suggest that GeckoGraph has potential as a teaching and learning aid for polymorphic type systems.
\end{abstract}

@@

 We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern Human-Computer Interaction (HCI) techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. 
 To evaluate its effectiveness, we conducted a large-scale user study comparing GeckoGraph with standard text-based type notation. To our knowledge, this represents the largest controlled user study on functional programming to date.
The results show no overall significant differences in task completion time or success rate. However, we observed a modest improvement in success rate among beginner participants. For the most challenging task, GeckoGraph improved success rates by 14.5\% for beginners and 11.2\% for less experienced users. These findings suggest that GeckoGraph has potential as a teaching and learning aid for polymorphic type systems.

@@

{\bf Structural typing} is often the primary approach for determining type equivalence and compatibility in functional programming languages (e.g., Elm and OCaml) and multi-paradigm languages (e.g., TypeScript). In contrast, nominal type systems are often predominant in Object-Oriented languages such as Java and C\# \cite{AbdelGawad2013-hh}. In nominal systems, two types are considered distinct even if they have identical structure but different names. In contrast, structural type systems consider types equivalent as long as their structures match, regardless of their names. This approach promotes flexibility, as types can be defined inline without introducing new names, and values are automatically valid for a type when they conform to the structure.  However, this lack of nominality can also make complex types harder to read and reason about. For instance, when encountering type errors, programmers can face two mismatched types with a complex structure, and understanding where exactly the root error is becomes a tedious task of string comparison (Fig. \ref{fig:structural}).

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/structural}
  \caption{\label{fig:structural} Example of a TypeScript structural type error. (Top) The function \texttt{f} expects a parameter with a string field \texttt{d}, but the argument {\tt x} provides a numeric {\tt d}. (Bottom) The error message is dreadful to read.}
\end{figure}

{\bf Algebraic data types} (ADTs) \cite{Milner1978-qz} allow developers to construct complex types from simpler ones using sum and product operations. A {\bf sum type}, often called a tagged union, $A + B$, represents a type that can be a value of A or a value of B. Conversely, a {\bf product type} $A \times B$ represents a tuple type composed of two values of both {\tt A} and {\tt B}. This formalization allows precise and composable type construction for common data structures such as lists, tuples, and result types.
For example, in Haskell, types like \texttt{data Either a b = Left a | Right b} and \texttt{data Maybe a = Nothing | Just a} provide basic building blocks for error handling, and the type \texttt{data List a = [] | a : List a } provides a succinct recursive definition for lists. This illustrates the expressive power of ADTs. However, in practice, type definitions can grow to be complex structures, for instance \texttt {Maybe (Either [Maybe a] (Either a String))} as programs grow, causing difficulties to interpret and maintain.

{\bf Parametric polymorphism} (often called generics) allows functions and data structures to operate uniformly across different types. For instance, in Haskell, the function \texttt{head :: [a] -> a}, taking a list as input and returning the first element, is a polymorphic function that works on lists of any type; integers, strings, and functions are all valid candidates for the generic variable \texttt{a}.  It provides a strong and principled form of abstraction, enabling code reuse without sacrificing type safety.

@@

{\bf Type inference}, or implicit typing, reduces the need for explicit type annotations by allowing the compiler to deduce types automatically.
Most modern languages employ some level of inference. For instance, Java’s var keyword (introduced in version 10) infers variable types at compile time based on their assigned values \cite{Java_Developers2023-an}, saving programmers from writing redundant type declarations, such as \textit {Car car = new Car()}.

@@

{\it To summarize}, the type systems of functional programming languages provide rigorous tools for specifying and enforcing program behavior. By encoding system constraints directly in types, they prevent many classes of runtime errors. However, this rigor introduces trade-offs: a steeper learning curve and less flexibility for exploratory or prototype development. In the following sections, we explore how GeckoGraph supports these type-system features by visualizing type-level information that is often implicit or difficult for humans to interpret.

@@

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, a type-class glyph is embedded in every type variable that requires this class. This means when displaying the type \texttt{(==) :: Eq a => 
a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in twice on both occurrences of the type variable \texttt{a}.

@@

\begin{figure}[h]
  \includegraphics[width=0.6\linewidth]{figures/Add3}
  \caption{\label{fig:add3} An implementation of function \texttt{add3} but the author missed an (+) from the correct implementation (.) ((+) .) (+). GeckGraph highlights the difference in arity and reveals that a binary function is needed on the right-hand side for the arity to match. }
\end{figure}

@@

\paragraph{Provide Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references, or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail, a specific query that is hard to execute with point and click over text based type annotation.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/Advanced}
  \caption{\label{fig:advanced}
  Advanced features of GeckoGraph. (A) Kind visualization. (B)  Qualified constraints. (C)  Multiple Parameter Type Classes. (D) Precise selection of sub-structures. }
\end{figure}

\section{Evaluation} \label{sec:evaluation}
To evaluate the effectiveness of GeckoGraph, we designed and conducted a gamified controlled experiment. The study, ``Zero to Hero", adopted a game-like experience. It contains 10 tasks (levels) of varying difficulty. At each level, participants are asked to implement a function named "\texttt{zeroToHero}" with the help of a limited list of provided functions. These functions are different at each level, and the target types of Zero and Hero vary at each level as well. The details of each level are provided in the Appendix \ref{levels}.

@@

We illustrate the tasks of the user study using levels 2 and 3 as examples. For level 2 (Fig. \ref{fig:level-example}), the players need to implement the function \texttt{ zeroToHero :: Zero a -> Hero b}. The available functions are \texttt{runZero::Zero a -> a}, \texttt{makeHero:: a -> Hero a}. A generic function \texttt{(\$)} is provided to improve the ergonomics of composing functions, but all tasks can be completed without the use of generic functions. The possible solutions and other details of the level can be found in Appendix \ref{levels}. To complete this level, players only need to use each of the provided functions once. One possible solution to this task would be {\tt zeroToHero z = makeHero (runZero z)}. The user illustrated in Fig. \ref{fig:level-example} did not reach a correct solution, as the type signature and GeckoGraph in {\it To Implement} panel do not match the one in {\it Your Result} panel.

@@

In level 3 (Fig. \ref{fig:level3}),  players need to implement the function {\tt zeroToHero :: Zero a -> Hero (a, a)}.  In this level, not all provided functions are useful. In particular,  players must discover that only \texttt{f1} and \texttt{f3} are necessary to produce the desired results. One implementation that satisfies the target type is \texttt{zeroTohero z = f3 (f1 z)}. Another possible implementation is {\tt zeroToHero z = f3 . f1 \$ z}, exercising the Haskell function composition idioms. Again, the user illustrated in Fig. \ref{fig:level3} did not reach a correct solution, as the type signature and GeckoGraphs in {\it To Implement} and {\it Your Result} do not match.

@@

Participants are assigned to one of two groups (A and B). The experiment uses a between-subjects design.  The tasks assigned to the two groups are the same. All participants receive both treatments (with and without GeckoGraph) during their runs, but in a different order.   {\bf Group A} is assisted by GeckoGraph on even levels and text-based type annotation on odd levels. {\it Group B} is assisted by text-based type annotation on even levels and GeckoGraph on odd levels. The number of participants in the two groups is counterbalanced.

@@

\subsection{Task Design} \label{subsection:task}
During the study, participants received 10 tasks. The tasks start off easy but gradually increase in difficulty.  In each task, a target type of the function \texttt{zeroToHero} is shown to the participants. Participants are provided with a list of available functions to implement the target function and to ensure it assumes the target type. This is to simulate the tasks of selecting useful functions from a library. Participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

\subsection{Measurements}
During the study, the time spent by participants on each task is recorded. We also record the resulting status of each level, whether it is a success or a failure. Before starting the study, participants nominate their level of Haskell experience on a four-level scale: beginner, familiar,  knowledgeable, and expert.  If a participant has completed all 10 levels (with the help of skipping), we invite the participant to complete a post-study survey. In it, we ask for them to rate how {\bf intuitive} the GeckoGraph design is, how {\bf distracting} they find GeckoGraph, and how {\bf helpful} GeckoGraph is, using a seven-point scale. In the end, we ask 2 open-ended questions: {\it What is your overall experience using GeckoGraph} and {\it What are your expectations on the potential applications of GeckoGraph}.

Data collection of the study was stopped after the planned cut-off period of 14 days. After the cut-off date, the ZeroToHero website is open-sourced and available for free evaluation\footnote{https://zerotohero.fly.dev}, and repeating our experiment. However, no further data was collected.

@@

\subsection{Success rate}
We saw that, overall, GeckoGraph provides a slightly higher success rate (96.88\%) than text-based type notation (94.62\%). This trend can be seen in every experienced group: beginner group (95.12\% vs. 92.68\%), familiar group (97.39\% vs. 93.34\%), knowledgeable group (96.82\% vs.96.06\%), and expert group (98.2\% vs.96.40\%). We saw the significance decrease as the user's experience increased. When performing a proportion test on each group, we see that the effect is most significant with the beginner group and reject the null hypothesis (z score = 2.0228, p-value = 0.0431), followed by the familiar group (z score = 1.7495, p-value = 0.0802). The knowledgeable group (z score = 1.0295, p-value = 0.3032) and the expert group (z score = 0.8660, p-value = 0.3756) show less significant differences between treatments.

When breaking down the result in each task (Fig. \ref{fig:success-rate}), we were able to reject the null hypothesis in task 10 of the beginner group (94.9\% with GecoGraph, and 80.4\% without GeckoGraph, p-value=0.0452) and task 10 of the familiar group (98.5\% with GecoGraph, and 87.3\% without GeckoGraph, p-value = 0.136).

Level 10 is the most cognitively demanding: the provided functions are the lengthiest, and the solution is challenging and deceptive. Programmers must use the provided rotate function {\it f2} to adjust the positions of the four components of the {\it Zero} value four times to get to the correct position. Completing the level requires applying the functions {\it f1}, {\it f2}, and {\it f3} in a nested sequence that is 8 layers deep. Consequently, programmers need to frequently switch contexts to recall the purpose of each function. We speculate that this level demands the mental flexibility typical of a Haskell expert to internalize these functions quickly. GeckoGraph helps less experienced programmers to achieve the same level of comprehension fluency by adding visual clarity.

@@

\subsection{Qualitative Feedback}
In their responses to the post-study survey, participants showed mostly neutral sentiments towards GeckoGraph's design and appeal. They find GeckoGraph moderately intuitive to use and that its appearance in the interface does not cause distraction.

For the question ``Do you find the GeckoGraph distracting", most of the participants rated a low score, with an average of 2.88 (Fig. \ref{fig:qualitative} left). For the question ``How intuitive do you find the GeckoGraph?", we saw a reverse correlation of experience (Fig. \ref{fig:qualitative} Middle): experts find the GeckoGraph most intuitive (5.07), followed by the knowledgeable group (4.87), and the familiar group (4.80). The beginner group found it to be the least intuitive but still rated it positively, albeit with a low score of 4.71. 
When answering the question ``How helpful do you find GeckoGraph in finding the solution?", the answer is more divided into different experience groups (Fig. \ref{fig:qualitative} right). It is slightly positive for beginners (mean=4.25) and slightly negative for the other groups, the familiar group (mean=3.86) and the knowledgeable group (mean=3.32). The expert group finds that GeckoGraph is moderately unhelpful (mean=2.95)

Two interesting trends emerge. First, experienced users are more likely to find GeckoGraph intuitive and easy to use. This aligns with our design decision to make GeckoGraph resemble text-based type annotation. Those familiar with text-based type annotation find GeckoGraph second nature to read. Second, the correlation reverses when it comes to perceived helpfulness. The more experienced the users are, the less likely they are to find GeckoGraph helpful. We speculate this pattern occurs because users who are experts in text-based type annotations naturally prefer reading text-based notation and tend to ignore GeckoGraph, diminishing its usefulness. Conversely, novice users are more likely to be drawn to GeckoGraph and find that its visualization offers more clarity and useful insights.

@@

\paragraph{Task design}
In our user study, most of the provided functions are very abstract. These functions are created by the authors solely for the gamified study. They are designed to be different from well-known Haskell functions to minimize the familiarity variable. They are also designed with an interest in being challenging and fun. These functions may not be the most representative of real-world Haskell programming.

@@

\subsection{Strengths}\label{sec:discussion:strengths}
Overall, our experimental results show no statistically significant difference in task completion time between conditions with and without GeckoGraph. However, we find that using GeckoGraph shows a significant albeit moderate effect on the task success rates for beginners. Particularly, in the hardest task (task 10), GeckoGraph improved the success rates across the two less experienced user groups by 14.5\% and 11.2\%, respectively. 
To interpret the observed advantages, we discuss the practical benefits that GeckoGraph offers to programmers.

@@

\section{Conclusion and Future Work}
In this paper, we propose GeckoGraph, a graphical notation for type annotations in functional programming languages. GeckoGraph aims to complement traditional text-based type notation by making reading, understanding, and comparing types easier. We conducted a large-scale human study using GeckoGraph, comparing it to text-based type notation. This is the largest user study on functional programming we are aware of. Although no significant effect was found on the overall time to complete each programming task, we found a modest effect on the success rate in the beginner group. In one of the more difficult programming tasks, we find that GeckoGraph significantly improves the success rate by 14.5\% for beginners and 11.2\% for less experienced programmers.
